Malletwand: the Pendulum as a Handheld Interface to Musical Timing

### 2024-02-08 paper/paper.tex
(WIP?) Paper

\usepackage{stfloats}

@@

\begin{figure*}[ptb]
  \centering
  \includegraphics[width=1\textwidth]{placeholder.png}
  \caption{The Malletwand controller with the prototype self-playing glockenspiel.}
  \label{fig:Banner}
\end{figure*}

@@

\begin{figure*}[h!]
  \centering
  \includegraphics[width=1\textwidth]{placeholder.png}
  \caption{(a) Illustration of the controller being operated by a participant. (b) Photo of the circuit board. (c) Block diagram of the electronic components involved.}
  \label{fig:WandConstruction}
\end{figure*}

The board is fabricated as one, in order to ease soldering and debugging as we build the prototype. After testing is complete, we cut the board at the thin connection (in Figure~\ref{fig:WandConstruction}b) and connect the two parts with thin wires (8 strands of 32 AWG). The handle part is glued inside a 3D-printed casing with the wires hanging down onto the bob part, which is inserted onto a holder made of polymeric foam glued in a translucent plastic sphere filled with down cotton. We add fishing wires to strengthen the connection and prevent the electric wires from excessive tension and fatigue.

@@

The calibration is triggered by a wireless radio signal. The operator is then expected to hold the handle and let the bob stay stationary for a short while. Signals from the accelerometer are continuously monitored until they reach a sufficiently low deviation throughout a fixed time window of 3 seconds. This yields an axis vector that acts as the reference for future motion estimation from sensor readings which is saved in the persistent flash storage of the microcontroller.

\subsubsection{Magnetometer calibration}
As our algorithm relies on the geomagnetometer to improve accuracy of pose estimation, the distortion of the ambient magnetic field should also be compensated. Since indoor environments contain prominent and variable magnetic interference, this calibration is done online during operation. This works by continuously collecting the readings of the magnetometer and finding an ellipsoid fit that tries to distort the collected data in the form of 3D vectors onto a unit sphere. We employ the adjusted least squares estimator introduced by \cite{Markovsky_2004_ALS} for the ellipsoid fitting algorithm.

\subsubsection{Motion signal processing}
Our IMU produces readings at a sampling rate of 100 Hz, and the filtering algorithms run at the same rate. At each time step, we first estimate the facing of the bob through a complementary filter combining signals from the gyroscope and the magnetometer. The result is turned into a rotational transform that cancels out the bob's rotation around its own axis. Sensor readings adjusted by this transform are then fed into an extended Kalman filter (EKF) that estimates the trajectory of the bob in 3D space, and more importantly, its current phase within the period of motion. This is the value that gets interpreted as the driving signal of musical time, and is transmitted to the actuator.

@@

\section{The ``Mallet''}

@@

The instrument is accompanied by two ``wand'' controllers and can operate in both individual and co-operative mode. Signals received from the controllers control the passing of musical time in a predetermined musical piece, similar to a musical box. Co-operative mode is entered by two controllers being operated simultaneously, where each controller corresponds to one voice (part) indicated by colours of the lights. The music only progresses when the motion of the two are roughly in synchronisation.

@@

@@

\begin{figure*}[h!]
  \centering
  \includegraphics[width=1\textwidth]{placeholder.png}
  \caption{(a) Overall photo of the instrument. (b) Close-up of a single sub-unit. (c) Block diagram of the electronic components involved.}
  \label{fig:MalletConstruction}
\end{figure*}

@@

We attach a demonstration video with this report. Please refer to it.

@@

\subsection{Extrapolations and Future Work}
As we have written in the Introduction section, we imagine that this interface is not limited to our glockenspiel instrument. Any device that plays music can employ a pendulum interface to invite participants into musical time. Hybrid formats, including projections and mappings, or musical devices remotely incited beyond kilometers, opens doors to tactility in a hybrid world.

@@

\section{Conclusion}
You have only 200 words remaining \char`\^ \char`\^

@@

\appendix
\section{Details of the Signal Processing Algorithms}
\label{appendix:signal-processing}

\subsection{Calibration}
\subsubsection{Orientation}
\subsubsection{Magnetometer}

\subsection{Estimation}
\subsubsection{Facing}
\subsubsection{Motion}

>>>>

\usepackage{stfloats}
\usepackage{amsmath}

@@

\begin{figure*}[h!]
  \centering
  \includegraphics[width=1\textwidth]{Mw_Banner.jpg}
  \caption{The Malletwand controllers with the prototype self-playing glockenspiel.}
  \label{fig:Banner}
\end{figure*}

@@

\begin{figure*}[t!]
  \centering
  \includegraphics[width=1\textwidth]{Mw_Wand.pdf}
  \caption{(a) Photo of the assembled circuit board before being cut. (b) Block diagram of the electronic components involved.}
  \label{fig:WandConstruction}
\end{figure*}

The board is fabricated as one, in order to ease soldering and debugging as we build the prototype. After testing is complete, we cut the board at the thin connection (in Figure~\ref{fig:WandConstruction}a) and connect the two parts with thin wires (8 strands of 32 AWG). The handle part is glued inside a 3D-printed casing with the wires hanging down onto the bob part, which is inserted onto a holder made of polymeric foam glued in a translucent plastic sphere filled with down cotton. We add fishing wires to strengthen the connection and prevent the electric wires from excessive tension and fatigue.

@@

The calibration is triggered by a wireless radio signal. The operator is then expected to hold the handle and let the bob stay stationary for a short while. Signals from the accelerometer are continuously monitored until they reach a sufficiently low deviation throughout a fixed time window of 3 seconds. This yields an axis vector that acts as the reference for future motion estimation from sensor readings which is saved in the persistent flash storage of the microcontroller. For details, please refer to Appendix~\ref{appendix:calib-ori}.

\subsubsection{Magnetometer calibration}
As our algorithm relies on the geomagnetometer to improve accuracy of pose estimation, the distortion of the ambient magnetic field should also be compensated. Since indoor environments contain prominent and variable magnetic interference, this calibration is done online during operation. This works by continuously collecting the readings of the magnetometer and finding an ellipsoid fit that tries to distort the collected data in the form of 3D vectors onto a unit sphere. We employ the adjusted least squares estimator introduced by \cite{Markovsky_2004_ALS} with complementary steps by \cite{Renaudin2010}.

\subsubsection{Motion signal processing}
Our IMU produces readings at a sampling rate of 100 Hz, and the filtering algorithms run at the same rate. At each time step, we first estimate the facing of the bob through a complementary filter, inspired by~\cite{Min_Complementary}, combining signals from the gyroscope and the magnetometer. The result is turned into a rotational transform that cancels out the bob's rotation around its own axis. Sensor readings adjusted by this transform are then fed into an extended Kalman filter (EKF) that estimates the trajectory of the bob in 3D space, and more importantly, its current phase within the period of motion. This is the value that gets interpreted as the driving signal of musical time, and is transmitted to the actuator.

@@

\section{The ``Mallets''}

@@

The instrument is accompanied by two ``wand'' controllers and can operate in both solo and co-operative mode. Signals received from the controllers drive the passing of musical time in a predetermined musical piece, similar to a musical box. Co-operative mode is entered by two controllers being operated simultaneously, where each controller corresponds to one voice (part) indicated by colours of the lights. The music only progresses when the motion of the two are roughly in synchronisation.

@@

\begin{figure*}[t!]
  \centering
  \includegraphics[width=1\textwidth]{Mw_Mallet.pdf}
  \caption{(a) Overall photo of the instrument. (b) Close-up of a single sub-unit. (c) Block diagram of the electronic components involved.}
  \label{fig:MalletConstruction}
\end{figure*}

@@

@@

Please refer to the attached video for a performance recording.

@@

\subsection{Extrapolations and Future Work}
As we have written in the Introduction section, we imagine that this interface not be limited to our glockenspiel instrument. Any device that plays music can employ a pendulum interface to invite participants into musical time. Hybrid formats, including projections and mappings, or musical devices remotely incited beyond kilometers, opens doors to tactility in a hybrid world.

@@

\section{Conclusion}
Malletwand is our first step in spreading the enjoyablility of music through a tactile medium. It is to our delight to have implemented the system and seeing it bear the unique approachability and adaptability to varied levels of musical expertise, adding to the state-of-the-art (pun intended) repertoire of physical-system controller interfaces. We realise that this interface has the potential to adapt to more instruments and scenarios, and put forward what we have envisaged. We share our work in the hope that our methods and insights can be useful to those who share the aspirations with us.

@@

\appendix
\section{Details of the Signal Processing Algorithms}
\label{appendix:signal-processing}
We try to outline the mathematical derivations of the algorithms. For details, please refer to the source code.

\subsection{Calibration}
\subsubsection{Orientation}
\label{appendix:calib-ori}
An accelerometer reading is a 3-dimensional vector. When a sufficiently low variation of readings is achieved during calibration, their average vector is normalized and taken as the reference axis $\mathbf{u}$. We find a quaternion $\mathbf{q}$ that rotates this vector to the $+z$ unit vector, $\mathbf{z}^+ = \begin{bmatrix} 0 & 0 & 1 \end{bmatrix}^\mathsf{T}$. This is achieved by setting $\mathbf{q}_1 = \begin{bmatrix} \mathbf{u} \times \mathbf{z^+} & (1 + \mathbf{u} \cdot \mathbf{z^+}) \end{bmatrix}$ and $\mathbf{q} = \mathbf{q}_1 \mathbin{\mathop{/}} \lvert \mathbf{q}_1 \rvert$.

This quaternion $\mathbf{q}$ transforms the reference frame of the accelerometer such that the $xy$ plane corresponds to the local ground, whose normal vector is the $z$ axis. The rotation specified by $\mathbf{q}$ is applied to future raw accelerometer readings $\tilde{\mathbf{a}}$ and gyroscope readings $\tilde{\mathbf{g}}$, assuming that their reference frames are aligned, to yield a carlibrated 6-axis reading: % , as well as the magnetometer readings after its own calibration has been carried out, in order to give an absolute orientation with reference to the ground (refer to the explanation below for further details).
\begin{equation}\label{eqn:calib-ori}
\begin{aligned}
\mathbf{a} &= \mathbf{q} \tilde{\mathbf{a}} \mathbf{q}^{-1}\text{,} \\
\mathbf{g} &= \mathbf{q} \tilde{\mathbf{g}} \mathbf{q}^{-1}\text{.} \\
\end{aligned}
\end{equation}

\subsubsection{Magnetometer}
\label{appendix:calib-mag}
This task can be formulated as follows: given a set of 3-dimensional vectors $\{ \tilde{\mathbf{m}}_{1 \dots n} \}$, find an affine transform $(\mathbf{Q}, \mathbf{o})$ such that the set of $\{\mathbf{Q} \tilde{\mathbf{m}}_i + \mathbf{o}\}$ best fits onto a unit sphere.

We follow the method of \cite{Renaudin2010}, employing the ALS estimator in \cite{Markovsky_2004_ALS} to find a fit $(\mathbf{A}, \mathbf{c})$ to the set of equations
$(\tilde{\mathbf{m}}_i - \mathbf{c})^\mathsf{T} \mathbf{A} (\tilde{\mathbf{m}}_i - \mathbf{c}) = 1$
and then calculating an inverse transform that converts this ellipsoid back to the unit sphere.

Since the ALS estimator only requires an accumulated matrix $\mathbf{\Psi}$ and not the individual $\tilde{\mathbf{m}}_i$, we are able to carry out the calculation online in constant rather than linear space.

\subsection{Estimation}
\subsubsection{Facing}
\label{appendix:est-facing}
Our complementary filter design follows that in~\cite{Min_Complementary}, namely, a filter satisfying the transfer equation
\begin{equation}
\begin{split}
& \hat{\theta} = \frac 1 {s^2 + k_\mathrm{p} s + k_\mathrm{i}} \left( s \dot\theta + (k_\mathrm{p} s + k_\mathrm{i}) \theta \right) \text{,} \\
& k_\mathrm{p} = \frac {\omega_0} {Q_\mathrm{P}},\qquad k_\mathrm{i} = \omega_0^2\text{.}
\end{split}
\end{equation}
This filter balances between a proportional signal $\theta$ and a derivative signal $\dot\theta$. Here we treat the 2-dimensional orientation on the $xy$ plane as the value being estimated. The calibrated magnetometer provides the $\theta$ signal, while the calibrated gyroscope provides $\dot\theta$.

The input signals are given by $\theta = -\mathop{\mathrm{arctan}} \frac {\mathbf{m}_y} {\mathbf{m}_x}$ and $\dot\theta = \mathrm{d}{\mathbf{g}_y} \mathbin{\mathop{/}} \mathrm{d}t$. Here $\mathbf{m}$ is the calibrated and aligned magnetometer reading,
\begin{equation}\label{eqn:est-facing}
\mathbf{m} = \mathbf{q} \ (\mathbf{Q}\tilde{\mathbf{m}} + \mathbf{o}) \ \mathbf{q}^{-1}\text{.}
\end{equation}
% , and $\mathbf{g}$ is the calibrated gyroscope reading $\mathbf{g} = \mathbf{q} \tilde{\mathbf{g}} \mathbf{q}^{-1}$.

We fix the pole quality factor as $Q_\mathrm{P} = 1 \mathbin{\mathop{/}} \sqrt{2}$ and manually and empirically tune the corner frequency at $\omega_0 = 24\ \text{rad/s} \approx 3.8\ \text{Hz}$. A slightly different value might be needed if the characteristics (models) of sensors change.

\subsubsection{Motion}
\label{appendix:est-motion}
In the extended Kalman filter we formulate the state vector as $\mathbf{x} = \begin{bmatrix} \omega & A & \theta & B & \phi \end{bmatrix}^\mathsf{T}$, where the observation $\mathbf{z} = \begin{bmatrix} A \cos \theta & B \cos (\theta + \phi) \end{bmatrix}^\mathsf{T}$ is the adjusted gyroscope readings projected onto the $xy$ plane. Varying $\theta$ over $[0, 2\pi)$ traces out an elliptic trajectory of this point and $\theta$ is the current phase within the period. $\omega$ is the rate of increment of $\theta$.

The prediction function is
\begin{equation}
f(\mathbf{x}) = \begin{bmatrix} \omega & A & \theta + \omega \Delta t & B & \phi \end{bmatrix}^\mathsf{T}\text{.}
\end{equation}
The process noise covariance matrix is set to be a diagonal matrix with the first entry prominently larger than the others, since $\omega$ is actively changed by the participant. If $A, \theta, B, \phi$ are assigned large process noise variances, the filter may frequently get stuck in a local optimum where all parameters oscillate, altogether stopping the steady progress of phase.

After the estimation of $A, \theta, B, \phi$ are obtained, we find the tilt angle of the major axis by $\delta = \frac 1 2 \mathop{\mathrm{arctan}} \frac {A^2 + B^2 \cos 2\phi} {B^2 \sin 2\phi}$, and calculate the difference $\theta - (\delta + \frac 1 4 \pi)$ to align the zero phase to the fastest-moving point.

\section{Source Repository}
All of our design files (printed circuit board assemblies, mechanical models) and source code are published at \newline \anonymize{\texttt{https://github.com/ayuusweetfish/Malletwand}}.
